Physics-based protein structure prediction and design using the confinement method

### 2013-01-11 Paper/paper.tex
changed paper 11_4

The calculation of free energy differences is of central importance in the simulation of biochemical
systems. It is particularly difficult to calculate between pairs of macromolecular
conformations as well as a computationally expensive task with existing
methods. In this work, the confinement approach is used to calculate absolute
free energies of biomolecular systems. The method provides two main advantages: it does not require a reaction coordinate or
transition path and it is fast to compute. The free energy calculated can be decomposed into a per
residue contribution in an 
approximate way. Per residue free energy allows us
to identify the reason behind conformational preferences in biomolecules. Through out the article we show its use in different
challenging modeling problems. In particular, we show its use in predicting the conformational
preference of chamaleon sequences (sequences with high sequence identity and different
folds). This sequence dependent conformational preferences and per residue free energy decomposition set the stage for the use of this 
method in protein design.

@@

In computational structural biology there are numerous cases where the free energy difference
between two well defined 
states plays an important role. Examples range from small to large conformational changes of proteins due to
ligand binding, change of pH or other conditions \cite{Meirovitch2007}, \cite{Chipot2007}, \cite{Jorgensen2004}. 
Free energy also plays an important role in the case of protein
folding. The ground breaking work of Christian B. Anfinsen and coworkers showed that the native
structures of small globular proteins have a unique, thermodynamically stable native structure with
their conformation at the global free energy minimum \cite{Anfinsen1973}. Regardless of the starting
point or even after unfolding 
it by changing different condition most proteins will go back to the native state after
reinstituting native like conditions. Often, the changes in free energy between different
conformations is small, but nonetheless crucial in favoring the native state
 since defects in protein folding may be
the molecular cause of a range of human genetic disorders. For example a misfolded protein known as
a
prion can misfold correctly folded proteins when entering a healthy organism.
\Alberto{I'm not sure I understand this here} \Arijit{Thus, Free energy can also act as a scale to identify between the misfolded and the native state of
the protein.} These facts further give free energy a special importance in structural biology.

However, the theoretical calculation of conformational free energy change becomes difficult if the
states of interest are very different \cite{Meirovitch2007}. In such scenario timescales involved in
such conformational transitions may
be beyond the sampling ability of classical molecular dynamics simulation.  Generally, in order to
calculate the free energy a pathway or reaction coordinate is needed with several intermediate structures.
Methods like umbrella sampling \cite{Torrie1977}, thermodynamic integration \cite{Tironi1994} along with classical MD can suffer from
overlap problem and require a large computational effort. The idea of a reaction coordinate became
even more
complicated in the case of protein folding.  As the free energy landscape for protein folding 
is rough \cite{Dill1997}, \cite{Dill2008}, the molecule 
can get trapped in an energy well during conformational sampling. Even with 
advanced methods like replica exchange molecular dynamics, the three dimensional structure may be
trapped in a local minima for a considerable time, giving rise to nonconverged distributions where
misfolded states might be selected as native like
% (give the example of human pin1ww domain from Benout Roux and Klaus Schultan).

The calculation of free energy has been successfully attempted by a number of groups \cite{Meirovitch2007}, 
\cite{Ytreberg2006} - \cite{Zheng2008}. In recent years methods like the Reference system method \cite{Ytreberg2006}, 
Decativated Morphing \cite{Park2008}, Orthogonal Space Random Work \cite{Zheng2008} or the Confinement 
Method \cite{Tyka2006}, \cite{Cecchini2009} to name a few have been used to calculate free energy
differences.
In this work we have applied the confinement method which was originally devoled by Tyka et. al. \cite{Tyka2006} and
Cecchini et. al. \cite{Cecchini2009}. This method relies on the fact that the free energy difference is a state function and 
thus it is independent of the sampling path. This approach uses a thermodynamic cycle where first, non-harmonic degrees of
freedom are removed by applying a series of restraints to the system. Then the free energy of the
remaining harmonic system is obtained using a normal mode calculation and combined with the
non-harmonic part to give the total absolute free energy.  Previously, this method was applied to
calculate the conformational free energy of a 16 amino acid residue peptide, known as BHP.  We have
first reproduced that known result. Then we have applied this method to larger proteins for the
first time. The free energy difference of pairs of conformations in proteins with similar sequence but completely
different fold was calculated. \Alberto{i made this part too long, have to summarize here and define
better inside text?-->} Encouraged by the result, we applied the method to help tackle the
problem of protein conformation ranking. We did this by selecting different targets from the
Critical Assessment of Structure Prediction (CASP) event. In this event, different research groups
try to find the 3D structure of a protein given a sequence in a blind way using their methodologies. Upon comparing
with the real structure it is often seen that several groups are able to generate better structures
than the ones they submit, but are not able to identify them. This is known as the ranking problem,
and this method is able to correctly rank structures in the majority of cases we have tried.
 Finally, we have been exploring the ability of working on a per residue basis to identify which
 residues are the major players in the stabilization/destabilization of a pair of conformations. Our
 initial experience makes us thinkg that this method might be helpful in identifying key residues
 for protein design.

@@

\subsection{The confinement method produces correct results in control experiments}

@@

\begin{table}
\begin{center}
\caption{The confinement method assigns a more favorable free energy to the experimentally
determined structure than to computer-generated predictions. For each target, we examined as many as
five predictions submitted by CASP participants. Positive $\Delta\Delta G$ values indicate that the
experimental structure is predicted to be more favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier & $\Delta\Delta G_{native \to best decoy}$ (kcal/mol) \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$ \\ \hline
     T0569       &    2KWY        &          $20.01 \pm 0.69$  \\ \hline
\end{tabular}
\end{center}

@@

\subsection{The confinement method correctly predicts the structural preferences of chameleon
sequences}

@@

The results of these calculations are presented in Figure~\ref{fig:orban}. The confinement method
identifies the correct structure for all five sequences. One of the hypothesis for fold switching
of chamelon sequence is that the structural transitions require states with diminished stability.
It is widely believed that if the free energy of the native state and the alternative state is within
 a range of $-5 Kcal/Mol$ then  it can quickly change fold when the stability of the native state 
decreases. The stability of the native state can decrease for a number reason ranging from chemical 
modification, breaking of disulphide bonds or mutations as in this case. The free energy differences, that
came out from our calculation ranges from
around 2.9 to 5.7 kcal/mol, which is consistent with the above hypothesis \cite{He2008}, \cite{{Alexander2009}}.

@@

\begin{table}
\begin{center}
\caption{Analysis of per residue free energy decomposition of GA95 and GB95 sequence reveal that certain residues prefer
either $3 \alpha$  or $4 \beta + \alpha$ structure. These observations and the reason behind such conformational preferences are listed in this
table.}
\label{table:orban_perresidue}
\begin{tabular}{| l | l | l |}
\hline
    Observation                                & $3 \alpha$ structure           &    $4 \beta + \alpha$ structure                       \\ \hline
    Leu-5 and Leu-7 favor                      & hydrophobic residues are       & oriented towards the                                  \\ 
    $4 \beta + \alpha$                         & exposed to the solvent         & hydrophobic core                                      \\ \hline
    Gln-11 favors $3 \alpha$                   & Gln-11 has H-bond with Glu-15  & no such H-bond                                        \\ \hline
    Glu-14 favors $4 \beta + \alpha$           & hydrophilic residue partially  & completely exposed                                    \\       
                                               & exposed to the solvent         & to the solvent                                        \\ \hline        
    Ala-26 favors  $4 \beta + \alpha$          & hydrophobic residue oriented   & oriented towards the                                   \\ 
                                               & towards surface                & hydrophobic core                                       \\ \hline
    Ile-49 favors $3 \alpha$                   & hydrophobic interaction        & exposed to the                                          \\
                                               & inside the protein             & solvent                                                \\ \hline 
    Val-54 favors $4 \beta + \alpha$           & hydrophobic residue exposed    & hydrophobic interaction                                \\
                                               & to the solvent                 & inside the protein                                      \\ \hline
    Glu-56 favors $3 \alpha$                   & hydrophilic residue partially  & completely exposed                                    \\
                                               & exposed to the solvent         & to the solvent                                        \\ \hline
    Phe-30 in GB sequence                      & exposed to the solvent         & oriented towards the                                  \\
    favors $4 \beta + \alpha$                  & in protein surface             & hydrophobic core                                      \\ \hline   
    Tyr-45 in GB Sequence                      & No H-bonding                   & H-bond with Asp-47                                    \\  
    favors $4 \beta + \alpha$                  &                                &                                                       \\ \hline
    Asp-47 in GA sequence                      & H-bond with Lys-50             & No H-bond, No Tyr-45                                  \\  
    favors $3 \alpha$                                 &                         & in GA sequence                                         \\ \hline
\end{tabular}
\end{center}
\end{table}

\begin{figure}
\includegraphics[width=3.5 in,height=4 in]{orban.pdf}
\label{fig:orban}
\caption{The confinement method correctly predicts the structural preferences of six chameleon
sequences. (A) The six sequences used in this study. (B) Each sequence adopts either a Protein
G-like fold (denoted GB) or a three-helix bundle fold (denoted GA). The relative free energies of
the two folds are reported for each sequence.}
\end{figure}

\begin{figure}
\includegraphics[width=6.5 in,height=5.4 in]{orban_full.pdf}
\label{fig:orban_full}
\caption{Orientation of residues of different sidechains in $4 \beta + \alpha$ and $3 \alpha$ form
A. Orientation of Hydrophobic residues Leu-5 and Leu-7.
B. Gln-11 has H-bond with Glu-15 in $3 \alpha$, which is absent in $4 \beta + \alpha$
C. Hydrophobic Ala-26
D. Hydrophobic Val-54.
E. Hydrophilic Glu-14.
F. The plot of free energy difference for GA95 and GB95 sequence. Positive peak stabilize $3 \alpha$ form and negative peak 
stabilize $4 \beta + \alpha$. 
G. Hydrophilic Glu-56 .
H. Orientation of Hydrophic Ile-49.
I. With GA95 sequence, Ile-49 have hydrophic interaction with both Leu-20 and Ile-30.
J. Larger Phe-30 in GB sequence, move towards the protein surface in $4 \beta + \alpha$.
K. Tyr-45 form H-bond with Asp-47 in $4 \beta + \alpha$.}  
\end{figure}

%\begin{table}
%\caption{The confinement method correctly reproduces the structural preferences of a series of
%designed peptides. Each sequence adopts either a Protein G-like fold ($4 \beta + \alpha$) or forms a
%helix bundle ($3 \alpha$).}
%\label{table:orban}
%\begin{center}
%\begin{tabular}{l l l}\hline
%Changes in & $\Delta G$ (kcal/mol)    & $\Delta G$ (kcal/mol)     \\
 %Sequences & ($3 \alpha$ fold)        & ($4 \beta + \alpha$ fold) \\ \hline
%GA95       & $0.000$                  & $2.901 \pm 0.475$         \\
%GB95       & $4.365 \pm 0.460$        & $0.000$                   \\
%GA88       & $0.000$                  & $ 3.941 \pm 0.517$        \\
%GB88       & $5.576 \pm 0.464$        & $0.000$                   \\
%GA98       &  &                           \\ \hline
%GB98       &                          &                           \\ \hline
%\end{tabular}
%\end{center}
%\end{table}

>>>>

Calculation of free energy differences is of central importance in the simulation of biochemical
systems. It is particularly difficult to calculate between pairs of macromolecular
conformations as well as a computationally expensive task with existing
methods. In this work, confinement approach is used to calculate absolute
free energies of biomolecular systems. This method provides two main advantages: it does not require a reaction coordinate or
transition path and it is fast to compute. Free energy calculated can be decomposed into a per
residue contribution in an 
approximate way. Per residue free energy allows us
to identify the reason behind conformational preferences in biomolecules. Through out the article we show its use in different
challenging modeling problems. In particular, we show its use in predicting the conformational
preference of chamaleon sequences (sequences with high sequence identity and different
folds). This sequence dependent conformational preferences and per residue free energy decomposition set the stage for the use of this 
method in protein design.

@@

In computational structural biology there are numerous cases where the free energy difference
between two well defined 
states plays an important role. Examples range from small to large conformational changes of proteins due to
ligand binding, change of pH or other conditions \cite{Meirovitch2007}, \cite{Chipot2007}, \cite{Jorgensen2004}. 
Free energy also plays an important role in the case of protein
folding. Ground breaking work of Christian B. Anfinsen and coworkers showed that the native
structures of small globular proteins have a unique, thermodynamically stable native structure with
their conformation at the global free energy minimum \cite{Anfinsen1973}. Regardless of the starting
point or even after unfolding 
it by changing different condition most proteins will go back to the native state after
reinstituting native like conditions. Often, the changes in free energy between different
conformations is small, but nonetheless crucial in favoring the native state
 since defects in protein folding may be
the molecular cause of a range of human genetic disorders. For example a misfolded protein known as
a
prion can misfold correctly folded proteins when entering a healthy organism.
\Alberto{I'm not sure I understand this here} \Arijit{Thus, Free energy can also act as a scale to identify between the misfolded and the native state of
the protein.} These facts further give free energy a special importance in structural biology.

However, the theoretical calculation of conformational free energy change becomes difficult if the
states of interest are very different \cite{Meirovitch2007}. In such scenario timescales involved in
such conformational transitions may
be beyond the sampling ability of classical molecular dynamics simulation.  Generally, in order to
calculate the free energy a pathway or reaction coordinate is needed with several intermediate structures.
Methods like umbrella sampling \cite{Torrie1977}, thermodynamic integration \cite{Tironi1994} along with classical MD can suffer from
overlap problem and requires a larger computational effort. The idea of a reaction coordinate became
even more
complicated in the case of protein folding.  As free energy landscape for protein folding 
is rough \cite{Dill1997}, \cite{Dill2008}, the molecule 
can get trapped in an energy well during conformational sampling. Even with 
advanced methods like replica exchange molecular dynamics, the three dimensional structure may be
trapped in a local minima for a considerable time, giving rise to nonconverged distributions where
misfolded states might be selected as native like
% (give the example of human pin1ww domain from Benout Roux and Klaus Schultan).

Calculation of free energy have been successfully attempted by a number of groups \cite{Meirovitch2007}, 
\cite{Ytreberg2006} - \cite{Zheng2008}. In recent years methods like the Reference system method \cite{Ytreberg2006}, 
Decativated Morphing \cite{Park2008}, Orthogonal Space Random Work \cite{Zheng2008} or the Confinement 
Method \cite{Tyka2006}, \cite{Cecchini2009} to name a few have been used to calculate free energy
differences.
In this work we have applied the confinement method which was originally devoled by Tyka et. al. \cite{Tyka2006} and
Cecchini et. al. \cite{Cecchini2009}. This method relies on the fact that the free energy difference is a state function and 
thus it is independent of the sampling path. This approach uses a thermodynamic cycle where first, non-harmonic degrees of
freedom are removed by applying a series of restraints to the system. Then the free energy of the
remaining harmonic system is obtained using a normal mode calculation and combined with the
non-harmonic part to give the total absolute free energy.  Previously, this method was applied to
calculate the conformational free energy of a 16 amino acid residue peptide, known as BHP.  We have
first reproduced that known result. Next we have applied this method to larger proteins for the
first time. The free energy difference of pairs of conformations in proteins with similar sequence but completely
different fold was calculated. \Alberto{i made this part too long, have to summarize here and define
better inside text?-->} Encouraged by the result, we applied the method to help tackle the
problem of protein conformation ranking. We did this by selecting different targets from the
Critical Assessment of Structure Prediction (CASP) event. In this event, different research groups
try to find the 3D structure of a protein given a sequence in a blind way using their methodologies. Upon comparing
with the real structure it is often seen that several groups are able to generate better structures
than the ones they submit, but are not able to identify them. This is known as the ranking problem,
and this method is able to correctly rank structures in the majority of cases we have tried.
 Finally, we have been exploring the ability of working on a per residue basis to identify which
 residues are the major players in the stabilization/destabilization of a pair of conformations. Our
 initial experience makes us thinkg that this method might be helpful in identifying key residues
 for protein design.

@@

\subsection{Confinement method produces correct results in control experiments}

@@

\begin{table}
\begin{center}
\caption{Confinement method assigns a more favorable free energy to the experimentally
determined structure than to computer-generated predictions. For each target, we examined as many as
five predictions submitted by CASP participants. Positive $\Delta\Delta G$ values indicate that the
experimental structure is predicted to be more favorable than any of the decoys.}
\label{table:casp_control}
\begin{tabular}{l l l}\hline
    CASP Target  & PDB Identifier & $\Delta\Delta G_{native \to best decoy}$ (kcal/mol) \\ \hline
     T0531       &    2KJX        &          $11.15 \pm 0.70$ \\ \hline
     T0538       &    2L09        &          $-3.00 \pm 0.47$ \\ \hline
     T0540       &    3MX7        &          $16.94 \pm 0.49$ \\ \hline
     T0559       &    2L01        &          $2.10 \pm 0.24$ \\ \hline
     T0560       &    2L02        &          $18.00 \pm 0.49$ \\ \hline
     T0569       &    2KWY        &          $20.01 \pm 0.69$  \\ \hline
\end{tabular}
\end{center}

@@

\subsection{Confinement method correctly predicts the structural preferences of chameleon
sequences}

@@

The results of these calculations are presented in Figure~\ref{fig:orban}. Confinement method
identifies the correct structure for all five sequences. One of the hypothesis for fold switching
of chamelon sequence is that the structural transitions require states with diminished stability.
It is widely believed that if the free energy of the native state and the alternative state is within
 a range of $5 Kcal/Mol$ then it can quickly change fold when the stability of the native state 
decreases. The stability of the native state can decrease for a number reason ranging from chemical 
modification, breaking of disulphide bonds or mutations as in this case. The free energy differences, that
came out from our calculation ranges from
around 3.5 to 5.0 kcal/mol, which is consistent with the above hypothesis \cite{He2008}, \cite{{Alexander2009}}.

@@

\begin{table}
\begin{center}
\caption{Analysis of per residue free energy decomposition of GA95 and GB95 sequence reveal that certain residues prefer
either $3 \alpha$  or $4 \beta + \alpha$ structure. These observations and the reason behind such conformational preferences are listed in this
table.}
\label{table:orban_perresidue}
\begin{tabular}{| l | l | l |}
\hline
    Information from per                                 & What is happening in           & What is happening in                      \\ 
    residue plot                                         & $3 \alpha$ conformation        & $4 \beta + \alpha$  conformation          \\ \hline
    Leu-7 favors                                         & hydrophobic residue is         & oriented towards the                      \\ 
    $4 \beta + \alpha$ (Figure~\ref{fig:orban_full}(A))  & exposed to the solvent         & hydrophobic core                          \\ \hline
    Gln-11 stabilizes $3 \alpha$                   & Gln-11 has H-bond                         & no such H-bond                           \\ 
    (Figure~\ref{fig:orban_full}(B))           &  with Glu-15                              &                                           \\ \hline 
    Glu-14 favors $4 \beta + \alpha$           & hydrophilic residue partially  & completely exposed                                    \\       
    (Figure~\ref{fig:orban_full}(E))           & exposed to the solvent         & to the solvent                                        \\ \hline        
    Ala-26 favors  $4 \beta + \alpha$          & hydrophobic residue            & oriented towards the                                   \\ 
     (Figure~\ref{fig:orban_full}(C))          & oriented towards surface       & hydrophobic core                                       \\ \hline
    Ile-49 favors $3 \alpha$                   & hydrophobic interaction        & sidechain exposed                                       \\
    (Figure~\ref{fig:orban_full}(H) and (I))   & inside the protein             & to the solvent                                          \\ \hline 
    Ile-49 favors GA more                      & more hydrophobic               &  unchanged                                              \\
    than GB                                    & interactions with residues     &                                                         \\ 
    (Figure~\ref{fig:orban_full}(H) and (I))  &  at 20, 30 in GA                &                                                         \\ \hline
    Val-54 favors $4 \beta + \alpha$           & hydrophobic residue            & hydrophobic interaction                                \\
    Figure~\ref{fig:orban_full}(D))             & exposed to the solvent        & inside the protein                                      \\ \hline
    Glu-56 favors $3 \alpha$                   & hydrophilic residue partially  & completely exposed                                    \\
    (Figure~\ref{fig:orban_full}(G))            & exposed to the solvent         & to the solvent                                        \\ \hline
    Phe-30 in GB sequence                      & exposed to the solvent         & oriented towards the                                  \\
    stabilizes $4 \beta + \alpha$                  & in protein surface             & hydrophobic core                                      \\ \hline   
    Tyr-45 in GB favors                      & No H-bonding                   & H-bond with Asp-47                                    \\  
    $4 \beta + \alpha$ (Figure~\ref{fig:orban_full}(K))&                 &                                                       \\ \hline
    Asp-47 in GA sequence                      & H-bond with Lys-50             & No H-bond, No Tyr-45                                  \\  
    favors $3 \alpha$                          &                         & in GA sequence                                         \\ \hline
\end{tabular}
\end{center}
\end{table}

\begin{figure}
\includegraphics[width=3.5 in,height=4 in]{orban.pdf}
\label{fig:orban}
\caption{Confinement method correctly predicts the structural preferences of six chameleon
sequences. (A) The six sequences used in this study. (B) Each sequence adopts either a Protein
G-like fold (denoted GB) or a three-helix bundle fold (denoted GA). The relative free energies of
the two folds are reported for each sequence.}
\end{figure}

\begin{figure}
\includegraphics[width=6.4 in,height=5.3 in]{orban_full.pdf}
\label{fig:orban_full}
\caption{Sidechain orientation of some important residues which stabilize/destabilize either of the $4 \beta + \alpha$ or $3 \alpha$ conformation.
The details are in Table~\ref{table:orban_perresidue} and the main text.  
A. Leu-7.
B. Gln-11 has H-bond with Glu-15 in $3 \alpha$, which is absent in $4 \beta + \alpha$
C. Ala-26
D. Val-54.
E. Glu-14.
F. plot of free energy difference for GA95 and GB95 sequence. Positive peak stabilize $3 \alpha$ form and negative peak 
stabilize $4 \beta + \alpha$. 
G. Glu-56 .
H. sidechain orientation of Hydrophic Ile-49, along with Leu-20 and Ile-30 with GA95 sequence
I. sidechain orientation of Hydrophic Ile-49, along with Ala-20 and Phe-30 with GB95 sequence
J. Tyr-45 forms H-bond with Asp-47 with GB95 sequence and $4 \beta + \alpha$ conformation
K. No H-bond possible between Leu-45 and Asp-47 with GB95 sequence and $4 \beta + \alpha$ conformation .}  
\end{figure}

%\begin{table}
%\caption{Confinement method correctly reproduces the structural preferences of a series of
%designed peptides. Each sequence adopts either a Protein G-like fold ($4 \beta + \alpha$) or forms a
%helix bundle ($3 \alpha$).}
%\label{table:orban}
%\begin{center}
%\begin{tabular}{l l l}\hline
%Changes in & $\Delta G$ (kcal/mol)    & $\Delta G$ (kcal/mol)     \\
 %Sequences & ($3 \alpha$ fold)        & ($4 \beta + \alpha$ fold) \\ \hline
%GA95       & $0.000$                  & $2.901 \pm 0.475$         \\
%GB95       & $4.365 \pm 0.460$        & $0.000$                   \\
%GA88       & $0.000$                  & $ 3.941 \pm 0.517$        \\
%GB88       & $5.576 \pm 0.464$        & $0.000$                   \\
%GA98       &  &                           \\ \hline
%GB98       &                          &                           \\ \hline
%\end{tabular}
%\end{center}
%\end{table}
